Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-12-08 Documents/Paper/ms.tex
log^2 volume ratio in DNS

where $L(\theta)$ is the likelihood function, and the expectation is taken
with respect to the ``thermal'' distributions which are intermediate
between the prior $\pi(\theta)$ and the posterior:
\begin{eqnarray}
p_\beta(\theta) \propto \pi(\theta)L(\theta)^\beta.\label{eq:annealing}
\end{eqnarray}
If the posterior distribution is a mixture of a slab and a spike,
then for some range of inverse temperatures $\beta$ the probability mass in
the slab and the spike will be comparable. At these temperatures the MCMC
method will be unable to mix and will give an incorrect estimate of
$\left<\log\left[L(\theta)\right]\right>_\beta$.

Nested Sampling \citep{skilling} overcomes these problems by working with an
alternative sequence of probability distributions, proportional to the prior
but with a hard constraint on the value of the likelihood:
\begin{eqnarray}
p_{l}(\theta) \propto \pi(\theta)\mathds{1}\left[L(\theta) > l\right]\label{eq:constrained_prior}
\end{eqnarray}
The normalising constants of these distributions are given by
\begin{eqnarray}
X(l) &=& \int \pi(\theta)\mathds{1}\left[L(\theta) > l\right] \, d\theta\label{eq:X}
\end{eqnarray}
which is the amount of prior mass that has likelihood greater than
$l$. As Nested Sampling proceeds, $X$ shrinks geometrically with time, so
the number of iterations required to compress from the slab to the spike is
of order $\log(V_{\rm slab}/V_{\rm spike})$. During this process Nested Sampling
visits states with likelihoods intermediate between those of the slab and the
spike, whereas no choice of $\beta$ in Equation~\ref{eq:annealing} will give
these states substantial probability.

\subsection{Diffusive Nested Sampling}
The main difficulty with Nested Sampling is coming up with methods to sample
the constrained prior distributions of Equation~\ref{eq:constrained_prior}.
Various implementations of Nested Sampling exist, with different strategies
for sampling Equation~\ref{eq:constrained_prior}, as well as other features
to infer whether the run was successful, or to improve the accuracy of
the marginal likelihood estimate
\citep[e.g.][]{multinest, importance, diamonds, radfriends}.
Diffusive Nested Sampling
\citep[DNS][]{dnest} is an alternative version of Nested Sampling that is
appropriate when MCMC is used to explore the parameter space\footnote{
A C++ implementation of DNS, called {\tt DNest3}, is available at
{\tt https://github.com/eggplantbren/DNest3}}.

>>>>

where $L(\theta)$ is the likelihood function, and the expectation is taken
with respect to the ``thermal'' distributions which are intermediate
between the prior $\pi(\theta)$ and the posterior:
\begin{eqnarray}
p_\beta(\theta) \propto \pi(\theta)L(\theta)^\beta.\label{eq:annealing}
\end{eqnarray}
If the posterior distribution is a mixture of a slab and a spike,
then for some range of inverse temperatures $\beta$ the probability mass in
the slab and the spike will be comparable. At these temperatures the MCMC
method will be unable to mix (the sampler will only visit the spike or the
slab, and not both) and will give an incorrect estimate of
$\left<\log\left[L(\theta)\right]\right>_\beta$.

Nested Sampling \citep{skilling} overcomes these problems by working with an
alternative sequence of probability distributions, proportional to the prior
but with a hard constraint on the value of the likelihood:
\begin{eqnarray}
p_{l}(\theta) \propto \pi(\theta)\mathds{1}\left[L(\theta) > l\right]\label{eq:constrained_prior}
\end{eqnarray}
The normalising constants of these distributions are given by
\begin{eqnarray}
X(l) &=& \int \pi(\theta)\mathds{1}\left[L(\theta) > l\right] \, d\theta\label{eq:X}
\end{eqnarray}
which is the amount of prior mass that has likelihood greater than
$l$. As Nested Sampling proceeds, $X$ shrinks geometrically with time, so
the number of iterations required to compress from the slab to the spike is
of order $\log(V_{\rm slab}/V_{\rm spike})$. In Diffusive Nested Sampling
(Section~\ref{sec:dns}), it takes of order $\log(V_{\rm slab}/V_{\rm spike})$
iterations to pass from the slab to the spike initially, and then
$\log(V_{\rm slab}/V_{\rm spike})^2$ time to revise this.
During this process Nested Sampling
visits states with likelihoods intermediate between those of the slab and the
spike, whereas no choice of $\beta$ in Equation~\ref{eq:annealing} will give
these states substantial probability.

\subsection{Diffusive Nested Sampling}\label{sec:dns}
The main difficulty with Nested Sampling is coming up with methods to sample
the constrained prior distributions of Equation~\ref{eq:constrained_prior}.
Various implementations of Nested Sampling exist, with different strategies
for sampling Equation~\ref{eq:constrained_prior}, as well as other features
to infer whether the run was successful, or to improve the accuracy of
the marginal likelihood estimate
\citep[e.g.][]{multinest, importance, diamonds, radfriends}.
Diffusive Nested Sampling
\citep[DNS][]{dnest} is an alternative version of Nested Sampling that is
appropriate when MCMC is used to explore the parameter space\footnote{
A C++ implementation of DNS, called {\tt DNest3}, is available at
{\tt https://github.com/eggplantbren/DNest3}}.
